---
title: "CARD/RGI Resistome Analysis (Chapter 3 Half II Extended)"
author: "Rebekah Lanning"
output: html_document
---

# CARD/RGI Resistome Analysis

NEON Epilithon Shotgun Metagenomics
Sites: LEWI, MART, POSE, WALK | Years: 2022-2023

Framing: One Health surveillance using CARD's comprehensive ARG catalog (Alcock et al. 2023) at the Drug Class level. Parallel WHO Critically Important Antimicrobials filtered subset (WHO 2019, 6th Revision) provides the direct surveillance-relevant answer to the chapter question. Inferential analyses (PERMANOVA, MMUPHin batch correction, MaAsLin2) supported by CARD's read-count-based abundance data.

## 1. Load Libraries

```{r setup-card, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(rstatix)
library(ggpubr)
library(vegan)
library(MMUPHin)
library(Maaslin2)
```

Helper function to collapse semi-colon separated multi-class entries into a short readable form
```{r}
shorten_drug_class <- function(x) {
  # Split semicolon-separated lists, keep first 2 components, mark "& others"
  sapply(x, function(s) {
    parts <- trimws(strsplit(s, ";")[[1]])
    if (length(parts) <= 1) return(s)
    if (length(parts) == 2) return(paste(parts, collapse = " + "))
    # 3+ parts: show first two and indicate more
    paste0(paste(parts[1:2], collapse = " + "), " (+", length(parts) - 2, ")")
  })
}
```

## 2. Set File Paths

```{r card-paths}
RGI_DIR        <- "~/Documents/NEON/March_final/CARD_0507/rgi_results/"
METADATA_FILE  <- "~/Documents/NEON/March_final/metadata/shotgun/filtered_metadata_metagenomes.tsv"
MANIFEST_FILE  <- "~/Documents/NEON/March_final/metadata/shotgun/shotgun_manifest.tsv"
FIG_DIR        <- "~/Documents/NEON/March_final/CARD_0507/figures/"
OUTPUT_DIR     <- "~/Documents/NEON/March_final/CARD_0507/"

dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(paste0(OUTPUT_DIR, "maaslin2_results/"),
           recursive = TRUE, showWarnings = FALSE)
```

## 3. Canonical Chapter Palettes

```{r card-palettes}
site_colors <- c(LEWI = "#E41A1C", MART = "#377EB8",
                 POSE = "#4DAF4A", WALK = "#984EA3")
year_reds   <- c(`2022` = "#FCC5C0", `2023` = "#A50F15")
```

## 4. Load Metadata

```{r card-metadata}
metadata <- read_tsv(METADATA_FILE, show_col_types = FALSE) %>%
  filter(year != 2021)

manifest <- read_tsv(MANIFEST_FILE, show_col_types = FALSE) %>%
  filter(year != 2021) %>%
  mutate(sample = gsub("_R1\\.fastq\\.gz$", "", R1))

meta <- manifest %>%
  left_join(metadata %>% select(dnaSampleID, sampleFilteredReadNumber,
                                instrument_model),
            by = "dnaSampleID") %>%
  mutate(site = factor(site, levels = c("LEWI", "MART", "POSE", "WALK")),
         year = factor(year, levels = c("2022", "2023"))) %>%
  select(sample, site, year, sampleFilteredReadNumber, instrument_model)

cat("Samples in metadata:", nrow(meta), "\n")
```

## 5. Read and Merge CARD Outputs

```{r card-merge}
card_files <- list.files(RGI_DIR,
                         pattern = "\\.gene_mapping_data\\.txt$",
                         full.names = TRUE)

cat("Number of CARD output files found:", length(card_files), "\n")

read_card_one <- function(filepath) {
  sample <- gsub("\\.gene_mapping_data\\.txt$", "", basename(filepath))
  read_tsv(filepath, show_col_types = FALSE) %>%
    mutate(sample = sample)
}

card_long <- map_dfr(card_files, read_card_one)

cat("Total CARD detections (across all samples):", nrow(card_long), "\n")
cat("Unique ARO terms detected:",
    n_distinct(card_long$`ARO Term`), "\n")
cat("Unique Drug Classes detected:",
    n_distinct(card_long$`Drug Class`), "\n")
```

## 6. Clean and Annotate

```{r card-clean}
card_clean <- card_long %>%
  rename(aro_term       = `ARO Term`,
         aro_accession  = `ARO Accession`,
         drug_class     = `Drug Class`,
         amr_gene_family = `AMR Gene Family`,
         resistance_mechanism = `Resistance Mechanism`,
         all_mapped_reads = `All Mapped Reads`,
         reference_length = `Reference Length`,
         avg_pct_coverage = `Average Percent Coverage`) %>%
  select(sample, aro_term, aro_accession, drug_class, amr_gene_family,
         resistance_mechanism, all_mapped_reads, reference_length,
         avg_pct_coverage) %>%
  left_join(meta, by = "sample") %>%
  filter(!is.na(site))
```

## 7. WHO Critically Important Antimicrobials Categorization

WHO 2019 (6th revision). https://www.who.int/publications/i/item/9789241515528

```{r who-card-categorize}
who_categorize_class <- function(drug_class) {
  case_when(
    grepl("carbapenem|penam|cephalosporin|cephamycin|monobactam|penem|beta-lactam",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("aminoglycoside|streptomycin|spectinomycin",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("fluoroquinolone|quinolone",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("macrolide|ketolide",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("glycopeptide|vancomycin|teicoplanin|lipoglycopeptide",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("polymyxin|colistin",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("rifamycin|rifampin",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("oxazolidinone|linezolid",
          drug_class, ignore.case = TRUE) ~ "Critically Important",
    grepl("trimethoprim|sulfonamide|sulfa",
          drug_class, ignore.case = TRUE) ~ "Highly Important",
    grepl("tetracycline|glycylcycline",
          drug_class, ignore.case = TRUE) ~ "Highly Important",
    grepl("phenicol|chloramphenicol",
          drug_class, ignore.case = TRUE) ~ "Highly Important",
    grepl("amphenicol",
          drug_class, ignore.case = TRUE) ~ "Highly Important",
    grepl("streptogramin|pleuromutilin|lincosamide",
          drug_class, ignore.case = TRUE) ~ "Highly Important",
    TRUE ~ "Important / Other"
  )
}

card_clean <- card_clean %>%
  mutate(who_category = who_categorize_class(drug_class))

cat("\nWHO category breakdown (CARD detections):\n")
print(card_clean %>% count(who_category, sort = TRUE))
```

## 8. Aggregate to Drug Class Level (Read Counts)

```{r card-drugclass-counts}
drugclass_counts <- card_clean %>%
  group_by(sample, drug_class) %>%
  summarize(reads = sum(all_mapped_reads), .groups = "drop")

drugclass_wide <- drugclass_counts %>%
  pivot_wider(names_from = drug_class, values_from = reads,
              values_fill = 0) %>%
  arrange(sample)

cat("Samples with CARD data:", nrow(drugclass_wide), "\n")
cat("Drug Classes detected:", ncol(drugclass_wide) - 1, "\n")
```


## 9. ARG count summaries

```{r}
arg_counts <- card_clean %>%
  group_by(sample, site, year, sampleFilteredReadNumber) %>%
  summarize(n_args_total = n(),
            n_aro_unique = n_distinct(aro_term),
            n_drugclass  = n_distinct(drug_class),
            n_who_ci_args = sum(who_category == "Critically Important"),
            total_reads_to_args = sum(all_mapped_reads),
            .groups = "drop") %>%
  mutate(args_per_million = n_args_total /
                            (sampleFilteredReadNumber / 1e6),
         who_ci_per_million = n_who_ci_args /
                              (sampleFilteredReadNumber / 1e6))

cat("\nARG count summary by site:\n")
print(arg_counts %>%
        group_by(site) %>%
        summarize(n_samples = n(),
                  mean_args = round(mean(n_args_total), 1),
                  median_args = median(n_args_total),
                  mean_who_ci = round(mean(n_who_ci_args), 1),
                  .groups = "drop"))

cat("\nARG count summary by year:\n")
print(arg_counts %>%
        group_by(year) %>%
        summarize(n_samples = n(),
                  mean_reads_million = round(mean(sampleFilteredReadNumber)/1e6, 1),
                  mean_args_raw = round(mean(n_args_total), 1),
                  mean_args_per_million = round(mean(args_per_million), 4),
                  .groups = "drop"))
```


## 10. Statistical tests 

```{r}
kw_site_count <- kruskal.test(n_args_total ~ site, data = arg_counts)
dunn_site_count <- arg_counts %>%
  rstatix::dunn_test(n_args_total ~ site, p.adjust.method = "BH")

cat("=== Site KW (CARD raw ARG count) ===\n")
print(kw_site_count)
cat("\n=== Dunn's BH ===\n")
print(dunn_site_count)

wilcox_year_count <- wilcox.test(args_per_million ~ year,
                                 data = arg_counts, exact = FALSE)
wilcox_year_who <- wilcox.test(who_ci_per_million ~ year,
                               data = arg_counts, exact = FALSE)

cat("\n=== Year Wilcoxon (depth-adjusted) ===\n")
print(wilcox_year_count)
cat("\n=== Year Wilcoxon (WHO CI subset) ===\n")
print(wilcox_year_who)
```


## 11. Panel A. ARG count by site

```{r}
dunn_sig <- dunn_site_count %>% filter(p.adj < 0.05)
p_count_site <- ggplot(arg_counts,
                       aes(x = site, y = n_args_total, fill = site)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = year), width = 0.15, size = 2.5, alpha = 0.9) +
  scale_fill_manual(values = site_colors, name = "Site") +
  scale_color_manual(values = c("2022" = "#333333", "2023" = "#888888"),
                     name = "Year") +
  labs(title    = "A. CARD ARG detections per sample by site",
       subtitle = sprintf("Kruskal-Wallis: chi-squared = %.2f, df = %d, p = %.4f",
                          unname(kw_site_count$statistic),
                          unname(kw_site_count$parameter),
                          kw_site_count$p.value),
       caption  = "*** p-adj<0.001, ** p-adj<0.01, * p-adj<0.05",
       x = "Site",
       y = "Number of CARD ARG detections per sample") +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9),
        plot.caption  = element_text(color = "grey40", size = 8, hjust = 0))
if (nrow(dunn_sig) > 0) {
  dunn_sig <- dunn_sig %>%
    mutate(stars = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01  ~ "**",
      p.adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    ))
  y_top  <- max(arg_counts$n_args_total) * 1.05
  y_step <- max(arg_counts$n_args_total) * 0.05
  p_count_site <- p_count_site +
    stat_pvalue_manual(dunn_sig,
                       label = "stars",
                       y.position = seq(y_top,
                                        y_top + y_step * (nrow(dunn_sig) - 1),
                                        length.out = nrow(dunn_sig)),
                       tip.length = 0.01, size = 5)
}
ggsave(paste0(FIG_DIR, "card_panelA_count_by_site.png"),
       p_count_site, width = 8, height = 6)
print(p_count_site)
```


## 12. Panel B. ARG count by year (depth-adjusted)

```{r}
year_stars <- case_when(
  wilcox_year_count$p.value < 0.001 ~ "***",
  wilcox_year_count$p.value < 0.01  ~ "**",
  wilcox_year_count$p.value < 0.05  ~ "*",
  TRUE                               ~ "ns"
)
p_count_year <- ggplot(arg_counts,
                       aes(x = year, y = args_per_million, fill = year)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = site), width = 0.15, size = 2.5, alpha = 0.9) +
  annotate("segment",
           x = 1, xend = 2,
           y = max(arg_counts$args_per_million) * 1.05,
           yend = max(arg_counts$args_per_million) * 1.05) +
  annotate("text",
           x = 1.5,
           y = max(arg_counts$args_per_million) * 1.10,
           label = year_stars, size = 6) +
  scale_fill_manual(values = year_reds, name = "Year") +
  scale_color_manual(values = site_colors, name = "Site") +
  labs(title    = "B. CARD ARG detection rate by year (depth-adjusted)",
       subtitle = sprintf("Wilcoxon rank-sum on depth-adjusted count: W = %.1f, p = %.4f",
                          unname(wilcox_year_count$statistic),
                          wilcox_year_count$p.value),
       caption  = paste("Counts normalized by sampleFilteredReadNumber.",
                        "\n*** p<0.001, ** p<0.01, * p<0.05, ns = not significant"),
       x = "Year",
       y = "CARD ARG detections per million filtered reads") +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9),
        plot.caption  = element_text(color = "grey40", size = 8, hjust = 0))
ggsave(paste0(FIG_DIR, "card_panelB_count_by_year.png"),
       p_count_year, width = 7, height = 6)
print(p_count_year)
```


## 13. Panel C. Top WHO critically important drug classes

```{r}
top_classes <- card_clean %>%
  group_by(drug_class, who_category) %>%
  summarize(n_detections = n(),
            n_samples = n_distinct(sample),
            n_unique_genes = n_distinct(aro_term),
            .groups = "drop") %>%
  arrange(desc(n_detections)) %>%
  slice_head(n = 20) %>%
  mutate(drug_class_short = shorten_drug_class(drug_class))

p_top_classes <- ggplot(top_classes,
                        aes(x = reorder(drug_class_short, n_detections),
                            y = n_detections,
                            fill = who_category)) +
  geom_col() +
  geom_text(aes(label = n_detections), hjust = -0.2, size = 3) +
  scale_fill_manual(values = c("Critically Important" = "#C7211C",
                                "Highly Important" = "#F39B7F",
                                "Important / Other" = "#A0A0A0"),
                    name = "WHO Importance") +
  coord_flip() +
  labs(title    = "C. Top 20 Drug Classes detected by CARD",
       subtitle = "Total detections across 39 samples, colored by WHO 2019 importance",
       caption  = paste("WHO 2019 (6th Revision): Critically Important Antimicrobials for Human Medicine.",
                        "\nMulti-class entries shortened: 'A + B (+N)' means A, B, and N additional classes."),
       x = NULL,
       y = "Number of detections") +
  theme_bw(base_size = 11) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9),
        plot.caption  = element_text(color = "grey40", size = 8, hjust = 0),
        axis.text.y   = element_text(size = 9),
        legend.position = "bottom") +
  expand_limits(y = max(top_classes$n_detections) * 1.15)

ggsave(paste0(FIG_DIR, "card_panelC_top_drug_classes.png"),
       p_top_classes, width = 10, height = 7)
print(p_top_classes)
```


## 14. Panel D. WHO CI prescence/abscence heatmap by sample

```{r}
who_ci_classes <- card_clean %>%
  filter(who_category == "Critically Important") %>%
  group_by(drug_class) %>%
  summarize(n_detections = n(), .groups = "drop") %>%
  arrange(desc(n_detections)) %>%
  slice_head(n = 15) %>%
  pull(drug_class)

heatmap_data <- card_clean %>%
  filter(drug_class %in% who_ci_classes) %>%
  group_by(sample, drug_class, site, year) %>%
  summarize(detected = 1, .groups = "drop") %>%
  complete(sample, drug_class, fill = list(detected = 0)) %>%
  left_join(meta %>% select(sample, site, year),
            by = "sample",
            suffix = c("", "_meta")) %>%
  mutate(site = coalesce(site, site_meta),
         year = coalesce(year, year_meta)) %>%
  select(-any_of(c("site_meta", "year_meta"))) %>%
  mutate(drug_class_short = shorten_drug_class(drug_class))

sample_order <- meta %>%
  arrange(site, year, sample) %>%
  pull(sample)

# Build short-name ordering matching the original drug_class order
class_order_short <- shorten_drug_class(who_ci_classes)

heatmap_data$sample <- factor(heatmap_data$sample, levels = sample_order)
heatmap_data$drug_class_short <- factor(heatmap_data$drug_class_short,
                                         levels = rev(class_order_short))

p_who_heat <- ggplot(heatmap_data,
                     aes(x = sample, y = drug_class_short,
                         fill = factor(detected))) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c(`0` = "grey92", `1` = "#C7211C"),
                    labels = c("Not detected", "Detected"),
                    name = NULL) +
  facet_grid(. ~ site, scales = "free_x", space = "free_x") +
  labs(title    = "D. WHO Critically Important Drug Class detections per sample",
       subtitle = "Top 15 most-detected WHO CI classes; samples grouped by site",
       caption  = paste("WHO 2019 (6th Revision): Critically Important Antimicrobials for Human Medicine.",
                        "\nMulti-class entries shortened: 'A + B (+N)' means A, B, and N additional classes."),
       x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40", size = 9),
        plot.caption    = element_text(color = "grey40", size = 8, hjust = 0),
        axis.text.x     = element_text(angle = 90, hjust = 1, vjust = 0.5,
                                       size = 7),
        axis.text.y     = element_text(size = 8),
        strip.text      = element_text(face = "bold", size = 10),
        legend.position = "bottom",
        panel.grid      = element_blank())

ggsave(paste0(FIG_DIR, "card_panelD_who_presence_heatmap.png"),
       p_who_heat, width = 13, height = 5.5)
print(p_who_heat)
```


## 15. Panel E. ResFinder vs CARD comparison

```{r}
resfinder_total <- 60
resfinder_who_ci <- 5

card_total <- nrow(card_clean)
card_who_ci <- card_clean %>%
  filter(who_category == "Critically Important") %>%
  nrow()

comparison_df <- tibble(
  database  = c("ResFinder", "CARD", "ResFinder", "CARD"),
  category  = c("All detections", "All detections",
                "WHO Critically Important", "WHO Critically Important"),
  count     = c(resfinder_total, card_total,
                resfinder_who_ci, card_who_ci)
) %>%
  mutate(database = factor(database, levels = c("ResFinder", "CARD")),
         category = factor(category,
                           levels = c("All detections",
                                      "WHO Critically Important")))

p_compare <- ggplot(comparison_df,
                    aes(x = category, y = count, fill = database)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c(ResFinder = "#7B3294", CARD = "#008837"),
                    name = "Database") +
  scale_y_log10(labels = scales::comma) +
  labs(title    = "E. ResFinder vs CARD detection counts",
       subtitle = "Same 39-sample dataset analyzed with two databases. Note log scale.",
       caption  = "ResFinder catalogs only horizontally-acquired clinical genes. CARD catalogs comprehensive resistome.",
       x = NULL,
       y = "Total detections (log scale)") +
  theme_bw(base_size = 11) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9),
        plot.caption  = element_text(color = "grey40", size = 8, hjust = 0))

ggsave(paste0(FIG_DIR, "card_panelE_resfinder_vs_card.png"),
       p_compare, width = 8, height = 5)
print(p_compare)
```


## 16. Beta-lactamase callout (parallel to ResFinder Panel D)

```{r}
beta_lactamases <- card_clean %>%
  filter(grepl("beta-lactam|carbapenem|penam|cephalosporin|monobactam",
               drug_class, ignore.case = TRUE)) %>%
  group_by(aro_term, amr_gene_family, drug_class) %>%
  summarize(n_detections   = n(),
            n_samples      = n_distinct(sample),
            total_reads    = sum(all_mapped_reads),
            mean_pct_cov   = round(mean(avg_pct_coverage), 1),
            sites          = paste(sort(unique(as.character(site))),
                                   collapse = ", "),
            years          = paste(sort(unique(as.character(year))),
                                   collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_detections))

cat("\n=== Beta-lactamase detections (CARD) ===\n")
cat("Total unique:", nrow(beta_lactamases), "\n\n")
print(beta_lactamases %>% slice_head(n = 20))

write_csv(beta_lactamases,
          paste0(OUTPUT_DIR, "card_beta_lactamases.csv"))
```


## 17. Summary tables for export 

```{r}
gene_summary <- card_clean %>%
  group_by(aro_term, drug_class, amr_gene_family, who_category) %>%
  summarize(n_detections = n(),
            n_samples    = n_distinct(sample),
            total_reads  = sum(all_mapped_reads),
            mean_pct_cov = round(mean(avg_pct_coverage), 2),
            sites        = paste(sort(unique(as.character(site))),
                                 collapse = ", "),
            years        = paste(sort(unique(as.character(year))),
                                 collapse = ", "),
            .groups = "drop") %>%
  arrange(who_category, desc(n_detections))

write_csv(gene_summary,
          paste0(OUTPUT_DIR, "card_gene_summary_who.csv"))

drugclass_summary <- card_clean %>%
  group_by(drug_class, who_category) %>%
  summarize(n_detections   = n(),
            n_unique_genes = n_distinct(aro_term),
            n_samples_with = n_distinct(sample),
            total_reads    = sum(all_mapped_reads),
            .groups = "drop") %>%
  arrange(who_category, desc(n_detections))

write_csv(drugclass_summary,
          paste0(OUTPUT_DIR, "card_drugclass_summary.csv"))

cat("\n=== Drug class summary (top by detection count) ===\n")
print(drugclass_summary %>% slice_head(n = 20))
```




