---
title: "HUMAnN3 Functional Potential Analysis"
output: html_document
date: "2026-05-01"
---

# NEON Epilithon Shotgun Metagenomics - Functional Potential
# Sites: LEWI, MART, POSE, WALK | Years: 2022-2023
# Comparisons: Site x Year (no seasonal data in shotgun dataset)

# HUMAnN3 v3.9 
# Nucleotide DB: ChocoPhlAn full
# Protein DB: UniRef90 EC-filtered (Diamond translated search)
# Pathway DB: MetaCyc (built into HUMAnN3)
# Comparisons: Site x Year (no seasonal data in shotgun dataset)

## 1. Load Libraries
```{r}
library(tidyverse)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
library(MMUPHin)
library(Maaslin2)
library(ggraph)
library(tidygraph)
```

## 2. Set File Paths
```{r}
PATHABUNDANCE_FILE <- "~/Documents/NEON/humann3_merged/merged_pathabundance_cpm.tsv"
PATHCOVERAGE_FILE  <- "~/Documents/NEON/humann3_merged/merged_pathcoverage.tsv"
GENEFAMILIES_FILE  <- "~/Documents/NEON/humann3_merged/merged_genefamilies_cpm.tsv"
METADATA_FILE      <- "~/Documents/NEON/March_final/metadata/shotgun/meta_with_seasons.csv"
OUTPUT_DIR         <- "~/Documents/NEON/March_final/humann3"
```

## 3. Load Metadata
```{r}
meta <- read_csv(METADATA_FILE) %>%
  mutate(
    site  = factor(site, levels = c("LEWI", "POSE", "WALK", "MART")),
    year  = factor(year)
  )
```

## 4. Load HUMAnN3 Pathway Abundance
```{r}
humann_path <- read_tsv(PATHABUNDANCE_FILE)
dim(humann_path)
head(humann_path[, 1:5])
```


## 5. Clean and filter pathway abundance table
```{r}
# Strip _cat_Abundance suffix from column names
colnames(humann_path) <- gsub("_cat_Abundance", "", colnames(humann_path))

# Rename first column to Pathway
colnames(humann_path)[1] <- "Pathway"

# Keep only unstratified rows (no | in pathway name)
# and remove UNMAPPED and UNINTEGRATED
humann_path_unstrat <- humann_path %>%
  filter(!str_detect(Pathway, "\\|"),
         !Pathway %in% c("UNMAPPED", "UNINTEGRATED"))

dim(humann_path_unstrat)
head(humann_path_unstrat[, 1:5])
```



## 6. Prepare matrix for analysis
```{r}
# Convert to matrix (samples as rows, pathways as columns)
path_mat <- humann_path_unstrat %>%
  column_to_rownames("Pathway") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id")

# Join with metadata
path_meta <- path_mat %>%
  left_join(meta %>% select(sample_id, site, year), by = "sample_id")

# Check all samples matched
nrow(path_meta)  # should be 40
sum(is.na(path_meta$site))  # should be 0
```

## 6b. MMUPHin batch correction for instrument effect
```{r}
# Prepare matrix for MMUPHin (features as rows, samples as columns)
path_for_correction <- humann_path_unstrat %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

# Convert CPM to proportions
path_proportions <- sweep(path_for_correction, 2, colSums(path_for_correction), "/")

# Prepare metadata aligned to matrix columns
meta_df <- meta %>% column_to_rownames("sample_id") %>% as.data.frame()
path_proportions <- path_proportions[, rownames(meta_df)]

# Run MMUPHin - same approach as Kraken2 script
fit_adjust <- adjust_batch(
  feature_abd = path_proportions,
  batch       = "instrument_model",
  covariates  = "site",
  data        = meta_df,
  control     = list(verbose = FALSE)
)

path_corrected <- fit_adjust$feature_abd_adj
dim(path_corrected)
```

## 7. PERMANOVA on batch-corrected pathway abundance (full marginal model)

### Hypothesis

H0: Site does not explain functional community composition after accounting for year

H0: Year does not explain functional community composition after accounting for site

HA: At least one term explains significant variation after accounting for the other

```{r}
# Transpose for vegan (samples as rows)
path_corrected_t <- t(path_corrected) %>% as.data.frame()

# Align metadata
meta_aligned <- meta_df[rownames(path_corrected_t), ]

# Bray-Curtis dissimilarity on corrected data
bc_dist_corrected <- vegdist(path_corrected_t, method = "bray")

# Full marginal PERMANOVA (each term tested after accounting for the other)
set.seed(123)
perm_full_marg <- adonis2(bc_dist_corrected ~ site + year,
                          data = meta_aligned,
                          permutations = 999,
                          by = "margin")
perm_full_marg
```

Interpretation:
- Site: R²=0.427, F=8.828, p=0.001 *** - site remains dominant even after controlling for year
- Year: R²=0.008, F=0.478, p=0.928 - year remains non-significant after controlling for site

For writing:
In the full marginal model, site remained a significant and dominant driver of epilithon 
functional community composition after controlling for year (PERMANOVA: R²=0.427, F=8.828, 
p=0.001). Year was not significant after controlling for site (R²=0.008, p=0.928), 
confirming that functional potential is structured primarily by location rather than 
interannual variation.


## 8. PCoA plot - batch-corrected functional potential
```{r}
site_colors <- c("LEWI" = "#E41A1C", "MART" = "#377EB8",
                 "POSE" = "#4DAF4A", "WALK" = "#984EA3")

# PCoA on corrected Bray-Curtis distance
pcoa_result <- cmdscale(bc_dist_corrected, eig = TRUE, k = 3)

# Variance explained
var_explained <- round(pcoa_result$eig / sum(pcoa_result$eig[pcoa_result$eig > 0]) * 100, 1)

# Build plot dataframe
pcoa_df <- data.frame(
  PC1       = pcoa_result$points[, 1],
  PC2       = pcoa_result$points[, 2],
  PC3       = pcoa_result$points[, 3],
  sample_id = rownames(pcoa_result$points)
) %>%
  left_join(meta, by = "sample_id")

# PC1 vs PC2
p_h3_12 <- ggplot(pcoa_df, aes(x = PC1, y = PC2, color = site)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(aes(group = site), level = 0.95, linetype = "dashed") +
  scale_color_manual(values = site_colors, name = "Site") +
  labs(title = "PC1 vs PC2",
       x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# PC1 vs PC3
p_h3_13 <- ggplot(pcoa_df, aes(x = PC1, y = PC3, color = site)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(aes(group = site), level = 0.95, linetype = "dashed") +
  scale_color_manual(values = site_colors, name = "Site") +
  labs(title = "PC1 vs PC3",
       x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC3 (", var_explained[3], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# Build marginal subtitle from perm_full_marg (created in fix #1, section 7)
overall_subtitle <- sprintf(
  "PERMANOVA (marginal): site R\u00b2 = %.3f, F = %.3f, p = %.3f | year R\u00b2 = %.3f, F = %.3f, p = %.3f",
  perm_full_marg$R2[1], perm_full_marg$F[1], perm_full_marg$`Pr(>F)`[1],
  perm_full_marg$R2[2], perm_full_marg$F[2], perm_full_marg$`Pr(>F)`[2]
)

# Combined
p_pcoa_humann <- (p_h3_12 | p_h3_13) +
  plot_annotation(
    title    = "Bray-Curtis PCoA - Functional Potential by Site",
    subtitle = overall_subtitle,
    caption  = paste0("40 samples | HUMAnN3 MetaCyc pathways | MMUPHin batch-corrected | ",
                      "PC1+PC2 = ", var_explained[1]+var_explained[2], "%"),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40", size = 9),
      plot.caption  = element_text(color = "grey40", size = 8, hjust = 0)
    )
  )

p_pcoa_humann
ggsave(paste0(OUTPUT_DIR, "/pcoa_humann3_corrected.png"),
       p_pcoa_humann, width = 14, height = 6, dpi = 300)
```

## 9. PERMANOVA by site overall

- how much does site explain when year isn't in the model at all?

### PERMANOVA by site hypothesis

H0: Epilithon functional community composition (MetaCyc pathway abundance) does not differ between sites (LEWI, MART, POSE, WALK)

HA: Functional community composition differs in at least one site

```{r}
set.seed(123)
perm_site <- adonis2(bc_dist_corrected ~ site,
                     data = meta_aligned,
                     permutations = 999)
print(perm_site)
```
Interpretation: 

- Functional community composition differs in at least one site - reject H0
- Site explains 42.7% of total variance in MetaCyc pathway abundance (R²=0.427, F=8.958, p=0.001)
- This is highly statistically significant (p=0.001)
- The F statistic of 8.958 indicates strong separation between sites

For writing:
PERMANOVA revealed that site was a significant driver of epilithon functional community 
composition (R²=0.427, F=8.958, p=0.001), explaining 42.7% of total variance in 
Bray-Curtis dissimilarity of MetaCyc pathway abundances.


## 10. PERMDISP by site

- within each site, how spread out are the samples around their group centroid in multivariate space? Do some sites have more variable functional communities than others?

### Hypothesis

H0: The dispersion (variability) of functional community composition does not differ between sites

HA: At least one site has significantly different dispersion from the others

```{r}
disp_site <- betadisper(bc_dist_corrected, meta_aligned$site)

set.seed(123)
perm_disp_site <- permutest(disp_site, permutations = 999)
print(perm_disp_site)
```

```{r}
# Visual
plot(disp_site, main = "PERMDISP - functional dispersion by site",
     col = c("#E41A1C", "#4DAF4A", "#984EA3", "#377EB8"))
```


```{r}
boxplot(disp_site, 
        col = c("#E41A1C", "#4DAF4A", "#984EA3", "#377EB8"),
        main = "Distance to centroid by site",
        ylab = "Distance to centroid")
```

- PERMDISP is significant (F=4.071, p=0.013) - reject H0
- Sites differ significantly in their within-group dispersion (variability)
- LEWI shows the highest dispersion - samples are most spread out from their centroid
- POSE shows the lowest dispersion - most internally consistent functional community
- WALK and MART are intermediate

- This means the PERMANOVA result (site p=0.001) reflects BOTH differences in 
  centroid location AND differences in within-site variability

For writing:
PERMDISP revealed significant differences in multivariate dispersion between sites 
(F=4.071, p=0.013). LEWI exhibited the greatest within-site variability in functional 
community composition, while POSE was the most internally consistent. The significant 
dispersion result suggests that the PERMANOVA site effect reflects both compositional 
differences between sites and differences in within-site functional variability. 
LEWI's elevated dispersion is consistent with its smaller sample size (n=6) and 
its geographic and hydrological distinctiveness among the four sites.


## 11. PERMANOVA by year overall

### Hypothesis

- how much does year explain when site isn't in the model at all?

H0: Epilithon functional community composition does not differ between years (2022, 2023)

HA: Functional community composition differs between years

```{r}
set.seed(123)
perm_year <- adonis2(bc_dist_corrected ~ year,
                     data = meta_aligned,
                     permutations = 999)
print(perm_year)
```
Interpretation: 

- Functional community composition does not differ between years.
- Fail to reject H0 - functional community composition does not differ between years
- Year explains only 0.7% of total variance (R²=0.007, F=0.281, p=0.996)
- p=0.996 is extremely non-significant - year has no detectable effect on functional composition
- This is consistent with the full marginal model result (year p=0.928)

For writing:
Year was not a significant driver of epilithon functional community composition in the 
univariate model (PERMANOVA: R²=0.007, F=0.281, p=0.996), explaining less than 1% of 
variance in Bray-Curtis dissimilarity of MetaCyc pathway abundances. This suggests that 
functional potential of epilithon biofilms was stable across the two sampling years 
after accounting for instrument batch effects.

## 12. Per-site year PCoA and PERMANOVA

### LEWI year analysis

- Does functional community composition within LEWI differ between 
  2022 and 2023? We already know from the overall PERMANOVA that year is a weak 
  driver across all sites - now we ask whether that interannual signal exists 
  specifically within LEWI.

#### Hypothesis

H0: Functional community composition at LEWI does not differ between 2022 and 2023

HA: Functional community composition at LEWI differs between years

```{r}
# Subset corrected matrix to LEWI samples
lewi_samples <- meta_aligned %>% filter(site == "LEWI") %>% rownames()
path_LEWI <- path_corrected_t[lewi_samples, ]
meta_LEWI <- meta_aligned[lewi_samples, ]

# Bray-Curtis
bc_LEWI <- vegdist(path_LEWI, method = "bray")

# PCoA
pcoa_LEWI <- cmdscale(bc_LEWI, eig = TRUE, k = 2)
var_LEWI <- round(pcoa_LEWI$eig / sum(pcoa_LEWI$eig[pcoa_LEWI$eig > 0]) * 100, 1)

# PERMANOVA ~ year
set.seed(123)
perm_LEWI_year <- adonis2(bc_LEWI ~ year, data = meta_LEWI, permutations = 999)
print(perm_LEWI_year)
```

```{r}
# Plot
year_reds <- c("2022" = "#FFAAAA", "2023" = "#990000")

pcoa_df_LEWI <- data.frame(
  PC1       = pcoa_LEWI$points[, 1],
  PC2       = pcoa_LEWI$points[, 2],
  sample_id = rownames(pcoa_LEWI$points)
) %>% left_join(meta, by = "sample_id")

p_LEWI_year <- ggplot(pcoa_df_LEWI, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = year), size = 4, alpha = 1) +
  stat_ellipse(aes(group = year, color = year),
               level = 0.95, linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = year_reds, name = "Year") +
  labs(title    = "LEWI - Functional PCoA by year",
       subtitle = sprintf("PERMANOVA: R\u00b2 = %.3f, F = %.3f, p = %.3f",
                   perm_LEWI_year$R2[1], perm_LEWI_year$F[1],
                   perm_LEWI_year$`Pr(>F)`[1]),
       x = paste0("PC1 (", var_LEWI[1], "%)"),
       y = paste0("PC2 (", var_LEWI[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", color = "black"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_LEWI_year)
```

# LEWI plot with convex hulls instead of ellipses
```{r}
# Calculate convex hulls for each year group
hulls_LEWI <- pcoa_df_LEWI %>%
  group_by(year) %>%
  slice(chull(PC1, PC2))

p_LEWI_year <- ggplot(pcoa_df_LEWI, aes(x = PC1, y = PC2)) +
  geom_polygon(data = hulls_LEWI,
               aes(fill = year, group = year),
               alpha = 0.15) +
  geom_point(aes(color = year), size = 4, alpha = 1) +
  scale_color_manual(values = year_reds, name = "Year") +
  scale_fill_manual(values = year_reds, name = "Year") +
  labs(title    = "LEWI - Functional PCoA by year",
       subtitle = sprintf("PERMANOVA: R\u00b2 = %.3f, F = %.3f, p = %.3f",
                   perm_LEWI_year$R2[1], perm_LEWI_year$F[1],
                   perm_LEWI_year$`Pr(>F)`[1]),
       caption  = "Convex hulls shown in place of confidence ellipses (n=3 per group)",
       x = paste0("PC1 (", var_LEWI[1], "%)"),
       y = paste0("PC2 (", var_LEWI[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", color = "black"),
        plot.subtitle = element_text(color = "grey40", size = 9),
        plot.caption  = element_text(color = "grey40", size = 8, hjust = 0))

print(p_LEWI_year)
```

Interpretation:
- Fail to reject H₀ - year does not significantly drive functional composition at LEWI
- 2022 shows wider spread than 2023 - consistent with the higher dispersion seen in PERMDISP
- Substantial overlap between years supports non-significant result
- Note: convex hulls used in place of ellipses due to n=3 per group
- Note: only 719 permutations possible due to small sample size

For writing:
Within LEWI, year was not a significant driver of functional community composition 
(PERMANOVA: R²=0.317, F=1.860, p=0.2). The 2022 samples showed greater spread in 
ordination space than 2023, consistent with LEWI's elevated overall dispersion. 
Convex hulls are shown in place of confidence ellipses due to the small sample 
size at this site (n=3 per year).


### MART year analysis

- Does functional community composition within MART differ between
  2022 and 2023? 

#### Hypothesis

H0: Functional community composition at MART does not differ between 2022 and 2023

HA: Functional community composition at MART differs between years

```{r}
# Subset corrected matrix to MART samples
mart_samples <- meta_aligned %>% filter(site == "MART") %>% rownames()
path_MART <- path_corrected_t[mart_samples, ]
meta_MART <- meta_aligned[mart_samples, ]

# Bray-Curtis
bc_MART <- vegdist(path_MART, method = "bray")

# PCoA
pcoa_MART <- cmdscale(bc_MART, eig = TRUE, k = 2)
var_MART <- round(pcoa_MART$eig / sum(pcoa_MART$eig[pcoa_MART$eig > 0]) * 100, 1)

# PERMANOVA ~ year
set.seed(123)
perm_MART_year <- adonis2(bc_MART ~ year, data = meta_MART, permutations = 999)
print(perm_MART_year)
```

```{r}
# Plot
year_blues <- c("2022" = "#9EC4E0", "2023" = "#1A3D5C")

pcoa_df_MART <- data.frame(
  PC1       = pcoa_MART$points[, 1],
  PC2       = pcoa_MART$points[, 2],
  sample_id = rownames(pcoa_MART$points)
) %>% left_join(meta, by = "sample_id")

p_MART_year <- ggplot(pcoa_df_MART, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = year), size = 4, alpha = 1) +
  stat_ellipse(aes(group = year, color = year),
               level = 0.95, linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = year_blues, name = "Year") +
  labs(title    = "MART - Functional PCoA by year",
       subtitle = sprintf("PERMANOVA: R\u00b2 = %.3f, F = %.3f, p = %.3f",
                   perm_MART_year$R2[1], perm_MART_year$F[1],
                   perm_MART_year$`Pr(>F)`[1]),
       x = paste0("PC1 (", var_MART[1], "%)"),
       y = paste0("PC2 (", var_MART[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", color = "black"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_MART_year)
```

Interpretation:
- Fail to reject H0 - year does not significantly drive functional composition at MART
- Year explains 9.6% of variance within MART (R²=0.096, F=1.381, p=0.194)
- Ellipses overlap substantially - no clear year-based clustering visible
- 2022 shows slightly wider spread than 2023

For writing:
Within MART, year was not a significant driver of functional community composition 
(PERMANOVA: R²=0.096, F=1.381, p=0.194), explaining less than 10% of variance in 
Bray-Curtis dissimilarity. The PCoA shows substantial overlap between years, 
suggesting functional stability at MART across the two sampling years.


### POSE year analysis

- Does functional community composition within POSE differ between
  2022 and 2023?

#### Hypothesis

H0: Functional community composition at POSE does not differ between 2022 and 2023

HA: Functional community composition at POSE differs between years

```{r}
# Subset corrected matrix to POSE samples
pose_samples <- meta_aligned %>% filter(site == "POSE") %>% rownames()
path_POSE <- path_corrected_t[pose_samples, ]
meta_POSE <- meta_aligned[pose_samples, ]

# Bray-Curtis
bc_POSE <- vegdist(path_POSE, method = "bray")

# PCoA
pcoa_POSE <- cmdscale(bc_POSE, eig = TRUE, k = 2)
var_POSE <- round(pcoa_POSE$eig / sum(pcoa_POSE$eig[pcoa_POSE$eig > 0]) * 100, 1)

# PERMANOVA ~ year
set.seed(123)
perm_POSE_year <- adonis2(bc_POSE ~ year, data = meta_POSE, permutations = 999)
print(perm_POSE_year)
```


```{r}
# Plot
year_greens <- c("2022" = "#B8E4B8", "2023" = "#1F7A1F")

pcoa_df_POSE <- data.frame(
  PC1       = pcoa_POSE$points[, 1],
  PC2       = pcoa_POSE$points[, 2],
  sample_id = rownames(pcoa_POSE$points)
) %>% left_join(meta, by = "sample_id")

p_POSE_year <- ggplot(pcoa_df_POSE, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = year), size = 4, alpha = 1) +
  stat_ellipse(aes(group = year, color = year),
               level = 0.95, linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = year_greens, name = "Year") +
  labs(title    = "POSE - Functional PCoA by year",
       subtitle = sprintf("PERMANOVA: R\u00b2 = %.3f, F = %.3f, p = %.3f",
                   perm_POSE_year$R2[1], perm_POSE_year$F[1],
                   perm_POSE_year$`Pr(>F)`[1]),
       x = paste0("PC1 (", var_POSE[1], "%)"),
       y = paste0("PC2 (", var_POSE[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", color = "black"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_POSE_year)
```

Interpretation:
- Reject H0 - year significantly drives functional composition at POSE
- Year explains 28.1% of variance within POSE (R²=0.281, F=3.129, p=0.009 **)
- Clear separation between 2022 and 2023 visible in PCoA - ellipses do not overlap
- This contrasts with LEWI and MART where year was not significant
- POSE is the only site showing significant interannual functional turnover

This explains why year was non-significant in the overall model - the POSE 
signal is being diluted by the other three sites where year had no effect.

For writing:
Within POSE, year was a significant driver of functional community composition 
(PERMANOVA: R²=0.281, F=3.129, p=0.009), explaining 28.1% of variance in 
Bray-Curtis dissimilarity. The PCoA shows clear separation between 2022 and 2023 
communities at POSE, with non-overlapping ellipses along PC1. This site-specific 
interannual signal was not detectable in the overall model (year p=0.928), 
suggesting that functional turnover at POSE was masked by stability at the 
other three sites. This finding warrants further investigation into 
environmental drivers of interannual functional change specific to POSE.


### WALK year analysis

- Does functional community composition within WALK differ between
  2022 and 2023?

#### Hypothesis

H0: Functional community composition at WALK does not differ between 2022 and 2023

HA: Functional community composition at WALK differs between years

```{r}
# Subset corrected matrix to WALK samples
walk_samples <- meta_aligned %>% filter(site == "WALK") %>% rownames()
path_WALK <- path_corrected_t[walk_samples, ]
meta_WALK <- meta_aligned[walk_samples, ]

# Bray-Curtis
bc_WALK <- vegdist(path_WALK, method = "bray")

# PCoA
pcoa_WALK <- cmdscale(bc_WALK, eig = TRUE, k = 2)
var_WALK <- round(pcoa_WALK$eig / sum(pcoa_WALK$eig[pcoa_WALK$eig > 0]) * 100, 1)

# PERMANOVA ~ year
set.seed(123)
perm_WALK_year <- adonis2(bc_WALK ~ year, data = meta_WALK, permutations = 999)
print(perm_WALK_year)
```

```{r}
# Plot
year_purples <- c("2022" = "#D4A8E0", "2023" = "#5B1A7A")

pcoa_df_WALK <- data.frame(
  PC1       = pcoa_WALK$points[, 1],
  PC2       = pcoa_WALK$points[, 2],
  sample_id = rownames(pcoa_WALK$points)
) %>% left_join(meta, by = "sample_id")

p_WALK_year <- ggplot(pcoa_df_WALK, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = year), size = 4, alpha = 1) +
  stat_ellipse(aes(group = year, color = year),
               level = 0.95, linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = year_purples, name = "Year") +
  labs(title    = "WALK - Functional PCoA by year",
       subtitle = sprintf("PERMANOVA: R\u00b2 = %.3f, F = %.3f, p = %.3f",
                   perm_WALK_year$R2[1], perm_WALK_year$F[1],
                   perm_WALK_year$`Pr(>F)`[1]),
       x = paste0("PC1 (", var_WALK[1], "%)"),
       y = paste0("PC2 (", var_WALK[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", color = "black"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_WALK_year)
```

Interpretation:
- Fail to reject H0 - year does not significantly drive functional composition at WALK
- Year explains 11.4% of variance within WALK (R²=0.114, F=0.904, p=0.497)
- Ellipses overlap substantially - no year-based clustering visible
- 2023 ellipse is much wider than 2022, suggesting higher within-year variability in 2023

For writing:
Within WALK, year was not a significant driver of functional community composition 
(PERMANOVA: R²=0.114, F=0.904, p=0.497). The PCoA shows substantial overlap 
between years, with 2023 samples showing wider dispersion than 2022. Functional 
community composition at WALK was stable across the two sampling years.

## 13. Combined 4-panel per-site year PCoA
```{r}
library(patchwork)

p_combined_sites <- (p_LEWI_year | p_MART_year) /
                    (p_POSE_year  | p_WALK_year) +
  plot_annotation(
    title    = "Functional Community Composition by Year - Per Site",
    subtitle = "Bray-Curtis PCoA | HUMAnN3 MetaCyc pathways | MMUPHin batch-corrected",
    caption  = "LEWI: convex hulls shown (n=3 per group) | Ellipses = 95% confidence",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40", size = 9),
      plot.caption  = element_text(color = "grey40", size = 8, hjust = 0)
    )
  )

p_combined_sites

ggsave(paste0(OUTPUT_DIR, "/pcoa_per_site_year_panel.png"),
       p_combined_sites, width = 14, height = 12, dpi = 300)
```

## 14. PERMDISP by year within POSE

### Hypothesis
H0: The dispersion of functional community composition does not differ between 2022 and 2023 within POSE

HA: Dispersion differs between years within POSE

```{r}
disp_POSE_year <- betadisper(bc_POSE, meta_POSE$year)

set.seed(123)
perm_disp_POSE_year <- permutest(disp_POSE_year, permutations = 999)
print(perm_disp_POSE_year)
```
```{r}
boxplot(disp_POSE_year,
        col = c("#B8E4B8", "#1F7A1F"),
        main = "POSE - Distance to centroid by year",
        ylab = "Distance to centroid")
```

Interpretation:
- Fail to reject H₀ - dispersion does not significantly differ between years at POSE 
  (F=2.275, p=0.147)
- This strengthens the PERMANOVA result - the significant year effect at POSE 
  reflects true compositional differences between centroids, not dispersion artifacts
- 2023 shows visually wider spread than 2022 in the boxplot, but this difference 
  is not statistically significant
- The POSE PERMANOVA result (p=0.009) is therefore robust

For writing:
PERMDISP confirmed that the significant year effect at POSE (PERMANOVA: p=0.009) 
reflects true differences in community composition between years rather than 
differences in within-group dispersion (PERMDISP: F=2.275, p=0.147). Although 
2023 samples showed slightly greater spread around their centroid than 2022, 
this difference was not statistically significant, supporting the interpretation 
that epilithon functional communities at POSE underwent genuine compositional 
shifts between 2022 and 2023.


# 1. Pathway Richness (Alpha Diversity)

Question: Does the number of detected MetaCyc pathways differ between sites and years? This measures within-sample functional 
diversity - how many unique metabolic pathways are present in each epilithon community.

Note: For HUMAnN3 functional data we use pathway richness (number of pathways with abundance > 0) rather than Shannon diversity, because pathway presence/absence is more biologically meaningful than evenness for functional potential.

```{r}
# Calculate pathway richness per sample
# Count pathways with CPM > 0 for each sample
richness_df <- humann_path_unstrat %>%
  column_to_rownames("Pathway") %>%
  summarise(across(everything(), ~ sum(. > 0))) %>%
  pivot_longer(everything(), 
               names_to = "sample_id", 
               values_to = "pathway_richness") %>%
  left_join(meta, by = "sample_id")

# Summary by site
cat("Pathway richness summary by site:\n")
print(richness_df %>%
  group_by(site) %>%
  summarise(mean   = round(mean(pathway_richness), 1),
            sd     = round(sd(pathway_richness), 1),
            min    = min(pathway_richness),
            max    = max(pathway_richness)))

# Summary by year
cat("\nPathway richness summary by year:\n")
print(richness_df %>%
  group_by(year) %>%
  summarise(mean   = round(mean(pathway_richness), 1),
            sd     = round(sd(pathway_richness), 1),
            min    = min(pathway_richness),
            max    = max(pathway_richness)))
```


### Statistical tests on pathway richness

H0 (site): Pathway richness does not differ between sites.
HA (site): Pathway richness differs in at least one site.

H0 (year): Pathway richness does not differ between 2022 and 2023.
HA (year): Pathway richness differs between years.

```{r}
library(rstatix)
library(ggpubr)

# Site comparison: 4 groups, KW + Dunn's with BH correction
cat("\n--- Kruskal-Wallis: pathway richness ~ site ---\n")
kw_site_richness <- kruskal.test(pathway_richness ~ site, data = richness_df)
print(kw_site_richness)

cat("\n--- Dunn's post-hoc with BH correction ---\n")
dunn_site_richness <- richness_df %>%
  rstatix::dunn_test(pathway_richness ~ site, p.adjust.method = "BH")
print(dunn_site_richness)

# Year comparison: 2 groups, KW + Wilcoxon
cat("\n--- Kruskal-Wallis: pathway richness ~ year ---\n")
kw_year_richness <- kruskal.test(pathway_richness ~ year, data = richness_df)
print(kw_year_richness)

cat("\n--- Wilcoxon rank-sum: pathway richness ~ year ---\n")
wilcox_year_richness <- wilcox.test(pathway_richness ~ year, data = richness_df)
print(wilcox_year_richness)
```


### Pathway richness boxplots

```{r}
# Site colors (in case this chunk is run standalone)
site_colors <- c("LEWI" = "#E41A1C", "MART" = "#377EB8",
                 "POSE" = "#4DAF4A", "WALK" = "#984EA3")
year_reds   <- c("2022" = "#FFAAAA", "2023" = "#990000")

# Site boxplot with Dunn's BH brackets
dunn_sig_site <- dunn_site_richness %>%
  filter(p.adj < 0.05) %>%
  mutate(label = paste0("p=", round(p.adj, 4)))

p_richness_site <- ggplot(richness_df,
                          aes(x = site, y = pathway_richness, fill = site)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = year), width = 0.15, size = 2.5) +
  stat_pvalue_manual(dunn_sig_site,
                     label = "label",
                     y.position = seq(max(richness_df$pathway_richness) + 15,
                                      max(richness_df$pathway_richness) + 15 + 10 * (nrow(dunn_sig_site) - 1),
                                      length.out = nrow(dunn_sig_site)),
                     tip.length = 0.01,
                     size = 3) +
  scale_fill_manual(values = site_colors, name = "Site") +
  scale_color_manual(values = c("2022" = "#333333", "2023" = "#888888"),
                     name = "Year") +
  labs(title    = "Pathway richness by site",
       subtitle = paste0("Kruskal-Wallis p = ",
                  round(kw_site_richness$p.value, 4),
                  " | Brackets = Dunn's post-hoc, BH corrected"),
       x = "Site", y = "Number of detected MetaCyc pathways") +
  theme_bw(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_richness_site)

# Year boxplot with Wilcoxon bracket
p_richness_year <- ggplot(richness_df,
                          aes(x = year, y = pathway_richness, fill = year)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = site), width = 0.15, size = 2.5) +
  stat_compare_means(comparisons = list(c("2022", "2023")),
                     method = "wilcox.test",
                     label = "p.format",
                     bracket.size = 0.5) +
  scale_fill_manual(values = year_reds, name = "Year") +
  scale_color_manual(values = site_colors, name = "Site") +
  labs(title    = "Pathway richness by year",
       subtitle = paste0("Kruskal-Wallis p=", round(kw_year_richness$p.value, 4),
                         " | Wilcoxon p=", round(wilcox_year_richness$p.value, 4),
                         " (bracket)"),
       x = "Year", y = "Number of detected MetaCyc pathways") +
  theme_bw(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40", size = 9))

print(p_richness_year)
```


```{r}
OUTPUT_DIR <- "~/Documents/NEON/March_final/humann3/figures/"
# (or whatever your existing OUTPUT_DIR is - check the top of the Rmd)

ggsave(paste0(OUTPUT_DIR, "pathway_richness_by_site.png"),
       p_richness_site,
       width = 8, height = 6)

ggsave(paste0(OUTPUT_DIR, "pathway_richness_by_year.png"),
       p_richness_year,
       width = 6, height = 6)

# Combined 2-panel for the chapter
library(patchwork)
p_richness_combined <- (p_richness_site | p_richness_year) +
  plot_layout(guides = "collect", widths = c(1.4, 1)) +
  plot_annotation(
    title    = "Functional alpha diversity (pathway richness)",
    subtitle = paste0("Wilcoxon rank-sum p=", round(wilcox_year_richness$p.value, 4)),
    theme = theme(plot.title    = element_text(face = "bold", size = 14),
                  plot.subtitle = element_text(color = "grey40", size = 9))
  )

ggsave(paste0(OUTPUT_DIR, "pathway_richness_combined.png"),
       p_richness_combined,
       width = 13, height = 6)

p_richness_combined
```


## 15. Differential Abundance - MaAsLin2

- For each MetaCyc pathway, is its abundance significantly different between groups?

- MaAsLin2 fits a linear model for each of the 656 pathways individually, testing whether pathway abundance is significantly associated with site or year. 

- Null and alternative hypotheses:

For site:

H0: The abundance of pathway X does not differ between sites (LEWI, MART, POSE, WALK) after controlling for year

HA: The abundance of pathway X differs significantly in at least one site compared to the reference site (LEWI)

For year:

H0: The abundance of pathway X does not differ between 2022 and 2023 after controlling for site

HA: The abundance of pathway X differs significantly between years

- Why q ≤ 0.25 and not 0.05?
With 656 pathways being tested, many false positives would occur at p=0.05. FDR correction controls the expected proportion of false discoveries. MaAsLin2's default threshold of 0.25 means up to 25% of significant results could be false positives - this is standard for exploratory microbiome research.

```{r}
maaslin_input <- as.data.frame(t(path_corrected))

maaslin_meta <- meta %>%
  column_to_rownames("sample_id") %>%
  select(site, year) %>%
  mutate(year = factor(year))

maaslin_input <- maaslin_input[rownames(maaslin_meta), ]

fit <- Maaslin2(
  input_data     = maaslin_input,
  input_metadata = maaslin_meta,
  output         = paste0(OUTPUT_DIR, "/maaslin2_results"),
  fixed_effects  = c("site", "year"),
  reference      = "site,LEWI;year,2022",
  normalization  = "NONE",
  transform      = "LOG",
  min_prevalence = 0.1,
  standardize    = FALSE,
  cores          = 1
)
```

```{r}
# How many significant associations?
sig <- read.table(paste0(OUTPUT_DIR, "/maaslin2_results/significant_results.tsv"),
                  header = TRUE, sep = "\t")
nrow(sig)
table(sig$metadata)

# Top 10 significant results
head(sig[order(sig$qval), ], 10)
```
Interpretation:

- 1,006 significant pathway associations detected (q ≤ 0.25)
- 874 site associations, 132 year associations
- Site drives far more differential pathway abundance than year
  - consistent with PERMANOVA results
- Top site-associated pathways include amino acid biosynthesis 
  and nitrifier denitrification - ecologically meaningful for 
  stream biofilm communities
- Negative coefficients = lower abundance relative to LEWI

```{r}
library(tidyverse)

# Read the FULL results table, not just significant
maaslin_all <- read_tsv("~/Documents/NEON/March_final/humann3/maaslin2_results/all_results.tsv")

# Same cleaning function as before
clean_pathway <- function(x) {
  x %>%
    str_replace("^[A-Z0-9.]+\\.\\.+", "") %>%
    str_replace_all("\\.+", " ") %>%
    str_replace_all("\\s+", " ") %>%
    str_trim() %>%
    str_to_sentence() %>%
    str_replace_all("\\bIi\\b", "II") %>%
    str_replace_all("\\bIii\\b", "III") %>%
    str_replace_all("\\bIv\\b", "IV") %>%
    str_replace_all("\\bVi\\b", "VI") %>%
    str_replace_all("\\bVii\\b", "VII") %>%
    str_replace_all("\\bL\\s", "L-") %>%
    str_replace_all("\\bD\\s", "D-")
}

heatmap_df <- maaslin_all %>%
  mutate(pathway_clean = clean_pathway(feature),
         signed_log_q  = -log10(qval) * sign(coef),
         signif_marker = ifelse(qval < 0.25, "*", ""),
         covariate_label = case_when(
           metadata == "site" ~ paste0(value, "\nvs. LEWI"),
           metadata == "year" ~ paste0(value, "\nvs. 2022"),
           TRUE ~ paste(metadata, value)
         ))

# Top 20 by best q-value across any covariate
top_pathways <- heatmap_df %>%
  group_by(pathway_clean) %>%
  summarize(min_q = min(qval, na.rm = TRUE)) %>%
  arrange(min_q) %>%
  slice_head(n = 20) %>%
  pull(pathway_clean)

heatmap_df <- heatmap_df %>%
  filter(pathway_clean %in% top_pathways)

# Order pathways by direction of strongest effect
pathway_order <- heatmap_df %>%
  group_by(pathway_clean) %>%
  summarize(strongest = signed_log_q[which.max(abs(signed_log_q))]) %>%
  arrange(strongest) %>%
  pull(pathway_clean)
heatmap_df$pathway_clean <- factor(heatmap_df$pathway_clean, levels = pathway_order)

col_order <- c("MART\nvs. LEWI", "POSE\nvs. LEWI", "WALK\nvs. LEWI", "2023\nvs. 2022")
heatmap_df$covariate_label <- factor(heatmap_df$covariate_label, levels = col_order)

fill_limit <- max(abs(heatmap_df$signed_log_q), na.rm = TRUE)

p_maaslin_heatmap <- ggplot(heatmap_df,
                            aes(x = covariate_label,
                                y = pathway_clean,
                                fill = signed_log_q)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(sprintf("%+.1f", signed_log_q), signif_marker),
                color = abs(signed_log_q) > fill_limit * 0.5),
            size = 2.8, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "white", "FALSE" = "black")) +
  scale_fill_gradient2(low      = "#2166AC",
                       mid      = "white",
                       high     = "#B2182B",
                       midpoint = 0,
                       limits   = c(-fill_limit, fill_limit),
                       name     = expression(-log[10](q) %*% sign(coef))) +
  labs(title    = "Differentially abundant MetaCyc pathways",
       subtitle = "MaAsLin2 | HUMAnN3 batch-corrected | top 20 pathways by minimum q-value",
       caption  = "Reference categories: site = LEWI, year = 2022. Asterisks mark q < 0.25.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x       = element_text(size = 10, lineheight = 0.9),
        axis.text.y       = element_text(size = 9),
        plot.title        = element_text(face = "bold", size = 13),
        plot.subtitle     = element_text(color = "grey40", size = 9),
        plot.caption      = element_text(color = "grey40", size = 8, hjust = 0),
        panel.grid        = element_blank())

print(p_maaslin_heatmap)

ggsave(paste0(OUTPUT_DIR, "maaslin_pathway_heatmap.png"),
       p_maaslin_heatmap,
       width = 11, height = 8.5)

```


```{r maaslin_panels, fig.width=11, fig.height=6.5}
library(tidyverse)
library(patchwork)

# -------- Paths --------
FIG_DIR <- "~/Documents/NEON/March_final/humann3/figures/"
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

# -------- Helper: name normalization --------
normalize_name <- function(x) {
  x %>%
    str_replace("^X", "") %>%
    str_replace_all("[^A-Za-z0-9]+", ".") %>%
    str_replace_all("\\.+", ".") %>%
    str_replace_all("^\\.|\\.$", "") %>%
    str_to_lower()
}

# -------- Helper: pathway name cleaning --------
clean_pathway <- function(x) {
  x <- str_replace(x, "^[A-Z0-9.]+\\.\\.+", "")
  x <- str_replace_all(x, "\\.+", " ")
  x <- str_replace_all(x, "\\s+", " ")
  x <- str_trim(x)
  x <- str_to_sentence(x)
  x <- gsub("\\b[Ii]v\\b",  "IV",  x)
  x <- gsub("\\b[Ii]ii\\b", "III", x)
  x <- gsub("\\b[Ii]i\\b",  "II",  x)
  x <- gsub("\\b[Vv]ii\\b", "VII", x)
  x <- gsub("\\b[Vv]i\\b",  "VI",  x)
  x <- gsub("\\bL\\s",      "L-",  x)
  x <- gsub("\\bD\\s",      "D-",  x)
  x <- gsub("\\bTca\\b",    "TCA", x)
  x <- gsub("\\bCtp\\b",    "CTP", x)
  x <- gsub("\\bAtp\\b",    "ATP", x)
  x
}

# -------- Read MaAsLin2 results --------
maaslin_all <- read_tsv("~/Documents/NEON/March_final/humann3/maaslin2_results/all_results.tsv")

# ============================================================
# PANEL B: Top 6 year-differential pathways
# ============================================================

year_top_features <- maaslin_all %>%
  filter(metadata == "year", qval < 0.25) %>%
  arrange(qval) %>%
  slice_head(n = 6) %>%
  pull(feature)

year_top_lookup <- tibble(
  feature_orig  = year_top_features,
  feature_norm  = normalize_name(year_top_features),
  pathway_clean = clean_pathway(year_top_features)
)

manual_short_names <- c(
  "Superpathway of cytosolic glycolysis plants pyruvate dehydrogenase and tca cycle" =
    "Glycolysis + PDH + TCA",
  "Superpathway of glycolysis pyruvate dehydrogenase tca and glyoxylate bypass" =
    "Glycolysis + PDH + TCA + glyoxylate",
  "Superpathway of hexitol degradation bacteria" =
    "Hexitol degradation",
  "Superpathway of menaquinol 8 biosynthesis II" =
    "Menaquinol-8 biosynthesis II",
  "Peptidoglycan biosynthesis v beta lactam resistance" =
    "Peptidoglycan biosynthesis V (beta-lactam resistance)",
  "Cis geranyl coa degradation" =
    "Cis-geranyl-CoA degradation"
)

year_top_lookup <- year_top_lookup %>%
  mutate(pathway_clean = ifelse(pathway_clean %in% names(manual_short_names),
                                manual_short_names[pathway_clean],
                                pathway_clean))

year_abundance_long <- path_corrected_t %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  pivot_longer(cols = -sample, names_to = "feature", values_to = "abundance") %>%
  mutate(feature_norm = normalize_name(feature)) %>%
  filter(feature_norm %in% year_top_lookup$feature_norm) %>%
  left_join(meta_aligned %>%
              as.data.frame() %>%
              rownames_to_column("sample"),
            by = "sample") %>%
  left_join(year_top_lookup, by = "feature_norm") %>%
  mutate(pathway_clean = factor(pathway_clean,
                                levels = year_top_lookup$pathway_clean))

year_star_y_pos <- year_abundance_long %>%
  group_by(pathway_clean) %>%
  summarize(y_pos = max(abundance) * 1.05, .groups = "drop")

year_stars_B <- maaslin_all %>%
  filter(metadata == "year") %>%
  mutate(feature_norm = normalize_name(feature)) %>%
  filter(feature_norm %in% year_top_lookup$feature_norm) %>%
  left_join(year_top_lookup, by = "feature_norm") %>%
  mutate(stars = case_when(
    qval < 0.001 ~ "***",
    qval < 0.01  ~ "**",
    qval < 0.05  ~ "*",
    qval < 0.25  ~ ".",
    TRUE         ~ "ns"
  )) %>%
  select(pathway_clean, stars) %>%
  left_join(year_star_y_pos, by = "pathway_clean") %>%
  mutate(pathway_clean = factor(pathway_clean,
                                levels = year_top_lookup$pathway_clean))

p_B <- ggplot(year_abundance_long,
              aes(x = year, y = abundance)) +
  geom_boxplot(fill = NA, color = "grey30", outlier.size = 0.8, linewidth = 0.5) +
  geom_jitter(aes(color = site), width = 0.15, size = 1.5, alpha = 0.9) +
  geom_text(data = year_stars_B,
            aes(x = 1.5, y = y_pos, label = stars),
            inherit.aes = FALSE,
            size = 5, color = "black") +
  facet_wrap(~ pathway_clean, scales = "free_y", ncol = 3) +
  scale_color_manual(values = site_colors, name = "Site") +
  labs(title    = "Top 6 year-differential pathways",
       subtitle = "Stars indicate BH-adjusted p-value, 2023 vs 2022 (2022 = reference)",
       caption  = "*** p-adj<0.001, ** p-adj<0.01, * p-adj<0.05, . p-adj<0.25, ns = not significant",
       x        = NULL,
       y        = "Relative abundance (batch-corrected)") +
  theme_bw(base_size = 10) +
  theme(plot.title      = element_text(face = "bold", size = 11),
        plot.subtitle   = element_text(color = "grey40", size = 9),
        plot.caption    = element_text(color = "grey40", size = 8, hjust = 0),
        strip.text      = element_text(size = 8),
        axis.text.x     = element_text(size = 9),
        legend.position = "right")

ggsave(paste0(FIG_DIR, "maaslin_panelB_pathways_by_year.png"),
       p_B, width = 11, height = 6.5)

print(p_B)

# ============================================================
# PANEL C: Mean abundance per site for top 8 site-differential pathways
# ============================================================

sig_pathways <- maaslin_all %>%
  filter(metadata == "site", qval < 0.25) %>%
  group_by(feature) %>%
  summarize(max_abs_coef = max(abs(coef)), .groups = "drop") %>%
  arrange(desc(max_abs_coef)) %>%
  slice_head(n = 8) %>%
  pull(feature)

sig_lookup <- tibble(
  feature_orig  = sig_pathways,
  feature_norm  = normalize_name(sig_pathways),
  pathway_clean = clean_pathway(sig_pathways)
)

panel_c_short_names <- c(
  "Mannosylglycerate biosynthesis i" =
    "Mannosylglycerate biosynthesis I",
  "Formaldehyde assimilation i serine pathway" =
    "Formaldehyde assimilation I (serine)",
  "Photosynthetic 3 hydroxybutanoate biosynthesis engineered" =
  "Photosynthetic 3-hydroxybutanoate",
  "Ceramide degradation by alpha oxidation" =
    "Ceramide degradation (alpha-oxidation)",
  "Pyruvate fermentation to butanoate" =
    "Pyruvate fermentation to butanoate",
  "L isoleucine biosynthesis IV" =
    "L-isoleucine biosynthesis IV",
  "L lysine biosynthesis III" =
    "L-lysine biosynthesis III",
  "Nitrifier denitrification" =
    "Nitrifier denitrification"
)

sig_lookup <- sig_lookup %>%
  mutate(pathway_clean = ifelse(pathway_clean %in% names(panel_c_short_names),
                                panel_c_short_names[pathway_clean],
                                pathway_clean))

site_panel_data <- path_corrected_t %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  pivot_longer(cols = -sample, names_to = "feature", values_to = "abundance") %>%
  mutate(feature_norm = normalize_name(feature)) %>%
  filter(feature_norm %in% sig_lookup$feature_norm) %>%
  left_join(meta_aligned %>%
              as.data.frame() %>%
              rownames_to_column("sample"),
            by = "sample") %>%
  left_join(sig_lookup, by = "feature_norm") %>%
  group_by(pathway_clean, site) %>%
  summarize(mean_ab = mean(abundance),
            se_ab   = sd(abundance) / sqrt(n()),
            .groups = "drop")

site_stars_simple <- maaslin_all %>%
  filter(metadata == "site") %>%
  mutate(feature_norm = normalize_name(feature)) %>%
  filter(feature_norm %in% sig_lookup$feature_norm) %>%
  left_join(sig_lookup, by = "feature_norm") %>%
  mutate(stars = case_when(
    qval < 0.001 ~ "***",
    qval < 0.01  ~ "**",
    qval < 0.05  ~ "*",
    qval < 0.25  ~ ".",
    TRUE         ~ ""
  )) %>%
  rename(site = value) %>%
  select(pathway_clean, site, stars)

site_panel_data <- site_panel_data %>%
  left_join(site_stars_simple, by = c("pathway_clean", "site")) %>%
  mutate(stars   = ifelse(is.na(stars), "", stars),
         star_y  = mean_ab + se_ab + max(mean_ab) * 0.10)

p_C <- ggplot(site_panel_data,
              aes(x = site, y = mean_ab, fill = site)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = mean_ab - se_ab, ymax = mean_ab + se_ab),
                width = 0.25, color = "grey30", linewidth = 0.4) +
  geom_text(aes(y = star_y, label = stars),
            size = 4, color = "black") +
  facet_wrap(~ pathway_clean, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = site_colors, name = "Site") +
  labs(title    = "Mean pathway abundance by site",
       subtitle = "Top 8 site-differential pathways | stars = BH-adjusted p-value vs LEWI",
       caption  = "*** p-adj<0.001, ** p-adj<0.01, * p-adj<0.05, . p-adj<0.25",
       x        = NULL,
       y        = "Mean relative abundance \u00b1 SE") +
  theme_bw(base_size = 10) +
  theme(plot.title      = element_text(face = "bold", size = 11),
        plot.subtitle   = element_text(color = "grey40", size = 9),
        plot.caption    = element_text(color = "grey40", size = 8, hjust = 0),
        strip.text      = element_text(size = 9),
        axis.text.x     = element_text(size = 9),
        legend.position = "none")

ggsave(paste0(FIG_DIR, "maaslin_panelC_site_means.png"),
       p_C, width = 12, height = 6.5)

print(p_C)
```
 


## 16. Sample Similarity Network - Functional Potential
# Question: Which samples share similar functional community profiles,
# and do samples cluster by site in network space?

# Method: Spearman correlation between samples based on pathway abundance,
# edges drawn between samples with correlation above threshold,
# nodes colored by site

```{r}
# ============================================================
# Sample similarity network (Spearman ρ ≥ 0.95)
# ============================================================
library(igraph)

# Spearman correlation between samples (columns of corrected matrix)
sample_cor <- cor(path_corrected, method = "spearman")

# Threshold: keep only strong positive correlations
adj <- sample_cor
adj[adj < 0.95] <- 0
diag(adj) <- 0  # remove self-loops

# Build undirected weighted graph
net <- graph_from_adjacency_matrix(adj, mode = "undirected", weighted = TRUE)

cat("Network: ", vcount(net), "nodes,", ecount(net), "edges\n")
```

```{r}
tg <- as_tbl_graph(net) %>%
  activate(nodes) %>%
  mutate(
    site = meta$site[match(name, meta$sample_id)],
    year = as.character(meta$year[match(name, meta$sample_id)])
  )

set.seed(123)
p_network <- ggraph(tg, layout = "fr") +
  geom_edge_link(aes(width = weight), color = "black", alpha = 1) +
  geom_node_point(aes(color = site, shape = year), size = 8) +
  scale_color_manual(values = site_colors, name = "Site") +
  scale_shape_manual(values = c("2022" = 16, "2023" = 17), name = "Year") +
  scale_edge_width(range = c(0.3, 1.5), guide = "none") +
  theme_graph(base_family = "sans") +
  labs(title    = "Sample Similarity Network - Functional Potential",
       subtitle = "Spearman \u03c1 \u2265 0.95 | nodes = samples | edges = high functional similarity") +
  theme(plot.title    = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "grey40", size = 9))

p_network
ggsave(paste0(OUTPUT_DIR, "/network_sample_similarity.png"),
       p_network, width = 8, height = 7, dpi = 300)

```
- MART (blue) clusters tightly on the left
- POSE + LEWI (green + red) cluster together on the right
- WALK (purple) sits in between, connecting both clusters
- Isolated nodes are samples with unique functional profiles - notably the LEWI samples

## 17. Pathway Co-occurrence Network

Question: Which MetaCyc pathways co-occur across samples suggesting functional coupling, and do functionally coupled pathways reflect known biological modules?

```{r}
# Use top 50 most variable pathways to keep network readable
path_var <- apply(path_corrected, 1, var)
top50 <- names(sort(path_var, decreasing = TRUE)[1:50])
path_top50 <- path_corrected[top50, ]

# Spearman correlation between pathways across samples
pathway_cor <- cor(t(path_top50), method = "spearman")
diag(pathway_cor) <- 0

# Threshold
pw_threshold <- 0.7
pathway_adj <- pathway_cor
pathway_adj[abs(pathway_adj) < pw_threshold] <- 0

# Build network using absolute correlations for layout
pw_net <- graph_from_adjacency_matrix(abs(pathway_adj),
                                      mode     = "undirected",
                                      weighted = TRUE,
                                      diag     = FALSE)

# Clean up pathway names for labels
V(pw_net)$label <- gsub(".*\\.\\.", "", V(pw_net)$name)
V(pw_net)$label <- gsub("\\.", " ", V(pw_net)$label)
V(pw_net)$label <- stringr::str_trunc(V(pw_net)$label, 30)

# Color by degree (connectivity)
V(pw_net)$degree <- degree(pw_net)

# Convert to tidygraph
pw_tg <- as_tbl_graph(pw_net)

set.seed(123)
p_pw_network <- ggraph(pw_tg, layout = "fr") +
  geom_edge_link(color = "black", alpha = 0.3, width = 0.5) +
  geom_node_point(aes(size = degree, color = degree)) +
  geom_node_text(aes(label = label), size = 2.5, repel = TRUE,
                 max.overlaps = 20) +
  scale_color_gradient(low = "#4DAF4A", high = "#E41A1C", name = "Connections") +
  scale_size(range = c(3, 10), name = "Connections", guide = "none") +
  theme_graph(base_family = "sans") +
  labs(title    = "Pathway Co-occurrence Network - Functional Potential",
       subtitle = "Top 50 most variable pathways | Spearman ρ ≥ 0.70 | node size = connectivity") +
  theme(plot.title    = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "grey40", size = 9))

p_pw_network
ggsave(paste0(OUTPUT_DIR, "/network_pathway_cooccurrence.png"),
       p_pw_network, width = 12, height = 10, dpi = 300)

```

Export network data from R
```{r}
# Export edge list
edges <- as_data_frame(pw_net, what = "edges")
write_csv(edges, paste0(OUTPUT_DIR, "/pathway_network_edges.csv"))

# Export node list  
nodes <- as_data_frame(pw_net, what = "vertices")
write_csv(nodes, paste0(OUTPUT_DIR, "/pathway_network_nodes.csv"))

write_graph(pw_net, 
            paste0(OUTPUT_DIR, "/pathway_network.graphml"), 
            format = "graphml")
```

```{r}
library(tidyverse)

# Build the data frame from your MaAsLin2 results
maaslin_counts <- tibble(
  metadata = c("Site", "Year"),
  n        = c(874, 132)
) %>%
  mutate(metadata = factor(metadata, levels = c("Site", "Year")))

p_counts <- ggplot(maaslin_counts, aes(x = metadata, y = n, fill = metadata)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), 
            vjust = -0.4, 
            size = 6, 
            fontface = "bold") +
  scale_fill_manual(values = c("Site" = "#2C5F8D", "Year" = "#B0B0B0"),
                    guide = "none") +
  scale_y_continuous(limits = c(0, 950), expand = c(0, 0)) +
  labs(title    = "Differentially abundant pathways by covariate",
       subtitle = "MaAsLin2 | q \u2264 0.25",
       x        = NULL,
       y        = "Number of significant associations") +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "grey40", size = 10),
        axis.text.x   = element_text(size = 12),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

print(p_counts)

ggsave(paste0(OUTPUT_DIR, "/maaslin_counts_bar.png"),
       p_counts, width = 5, height = 4)
```
